---
title: "Eurostat 2019 Analysis"
author: "Jason Ballantyne"
date: "11/10/2021"
output:
  pdf_document: default
  word_document: default
geometry: left=3cm,right=3cm,top=2cm,bottom=2cm
---


```{r, global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, tidy.opts=list(width.cutoff=60), tidy=TRUE)
options(warn = -1)
```

# Task 1

## 1.

```{r}
# Import the crime data
CrimeStats <- read.csv(file = "C:\\Users\\jason\\OneDrive - University College Dublin\\Documents\\MSc Computer Science\\Final Semester\\STAT40730 - Data Prog with R\\Assignment 1\\EurostatCrime2019.csv", row.names = 1)
head(CrimeStats)
```

## 2.

```{r}
# Getting the size (number of rows and columns)

# Getting the number of rows
nrow(CrimeStats)

# Getting the number of columns
ncol(CrimeStats)


# An alternative method for checking the number of rows and columns
# dim(CrimeStats)

# Getting the structure of the dataset
str(CrimeStats)
```
As per the above output:

* The number of rows are `r nrow(CrimeStats)`
* The number of columns are `r ncol(CrimeStats)`
* The structure of the dataset is a data frame.


## 3.
## (i)

```{r}
# Removing the columns Rape and Sexual Assault
CrimeStats <- CrimeStats[, !(names(CrimeStats) %in% c("Rape", "Sexual.assault"))]
```

## (ii)
```{r}
# Removing the columns Theft, 
# Theft.of.a.motorized.land.vehicle, 
# Burglary, Burglary.of.private.residential.premises 

CrimeStats <- CrimeStats[, !(names(CrimeStats) 
                             %in% c("Theft", 
                                    "Theft.of.a.motorized.land.vehicle", 
                                    "Burglary", 
                                    "Burglary.of.private.residential.premises"))]

# Checking that all the columns from (i) & (ii) have been sucessfully removed
head(CrimeStats)
```

## (iii)
```{r}
# Adding a column that contains the overall record of offences for each country

CrimeStats$Overall.Offences <- rowSums(Filter(is.numeric, CrimeStats), na.rm = TRUE)

# Checking that the new column with a record of the overall offences has been added

head(CrimeStats)
```

## 4.

```{r}
# Listing the countries that contain missing data

# We will firstly add any row with an NA to the dataframe CountryList
# Then we will print the list of countries that contain missing data

CountryList <- CrimeStats[rowSums(is.na(CrimeStats)) > 0,]
rownames(CountryList)
```

The above output tells us that the countries that contain missing data include:

* Belgium
* Bosnia and Herzegovina
* Denmark
* England and Wales
* Estonia
* France
* Hungary
* Iceland
* Liechtenstein
* Netherlands
* North Macedonia
* Northern Ireland (UK)
* Norway
* Poland
* Portugal
* Scotland
* Slovakia
* Sweden
* Turkey

## 5.

```{r}
# Removing the countries with missing data
CrimeStats <- na.omit(CrimeStats)

# An alternative method to removing the countries with missing data
# CrimeStats <- CrimeStats[rowSums(is.na(CrimeStats)) == 0, ]

# Summing the number of NA's to prove that missing data 
# has been removed from the dataframe
sum(is.na(CrimeStats))

```
Based on the above, there is `r sum(is.na(CrimeStats))` data missing, which shows we have successfully removed the countries with missing data from the dataframe.

## 6. 

```{r}
# Checking how many observations and variables are in the new dataframe

# Checking the number of observations
nrow(CrimeStats)

# Checking the number of variables
ncol(CrimeStats)

# Alternative method for checking the observations and variables
# dim(CrimeStats)
```


* There are `r nrow(CrimeStats)` observations in the new dataset
* There are `r ncol(CrimeStats)` variables in the new dataset


# Task 2

## 1.

```{r}

# Creating a new dataframe that contains all crimes in Ireland for 2019

IrelandCrimes <- CrimeStats["Ireland", 
                            c("Intentional.homicide", 
                              "Attempted.intentional.homicide",
                              "Assault", "Kidnapping", 
                              "Sexual.violence","Robbery", 
                              "Unlawful.acts.involving.controlled.drugs.or.precursors" )]

# Sorting the dataframe in Descending order
IrelandCrimes <- sort(IrelandCrimes, decreasing = TRUE)

# Extracting the first three columns to find the 3 most common crimes in Ireland
IrelandCrimes[,1:3]

```

The 3 most common crimes in Ireland in 2019 based on the above output were:

1. Unlawful acts involving controlled drugs or precursors
2. Assault
3. Sexual violence


## 2.


```{r}
# Creating a new dataframe that contains Assault and Overall.offences 
# figures in Ireland for 2019
Ireland_Assault <- CrimeStats["Ireland", c("Assault", "Overall.Offences")]

# Dataframe that calculates the proportion/percent of the 
# overall crimes that was due to assault 
Assault_Percent <- (Ireland_Assault$Assault/Ireland_Assault$Overall.Offences) * 100
Assault_Percent
```

Based on the above output, `r Assault_Percent`% of all crime in Ireland in 2019 was due to assault.

## 3.

```{r}
# Creating a new dataframe that contains the row with the maximum/highest 
# record of kidnapping in 2019
Kidnapping_Max <- CrimeStats[which.max(CrimeStats$Kidnapping),]

# Extracting the first column to display which country had the 
# highest record of kidnapping in 2019
rownames(Kidnapping_Max)
```

* Based on the output above, `r rownames(Kidnapping_Max)` had the highest record of kidnapping in 2019.
 
 
## 4.

```{r}
# Creating a new dataframe that contains the row with the lowest overall 
# recorded offences in 2019
Lowest_Offence <- CrimeStats[which.min(CrimeStats$Overall.Offences),]

# Extracting the first column to display which country had the 
# lowest overall record of offences in 2019
rownames(Lowest_Offence)
```
* Based on the output above, `r rownames(Lowest_Offence)` had the 
lowest overall record of offences in 2019.


## 5.

```{r}
# load packages
library(ggplot2)
library(ggthemes)
library(dplyr)

# Creating a plot to display the relationship between robbery and unlawful acts
# Plot was made to look "nice" through the use of ggplot, ggthemes and dplyr
# A regression line was added to show the trend line of my scatter plot at a glance
# Axis labels, a title, subtitle, and coloured shapes have all been included

CrimeStats %>%
  ggplot(aes(x = Robbery, y = 
               Unlawful.acts.involving.controlled.drugs.or.precursors)) +
  geom_point(size=3,shape=21, color="blue") +
  geom_smooth(formula = y ~ x, 
              method=lm, se=FALSE, linetype="dashed", color="darkred") +
  labs(title = "Robbery and Unlawful Acts",
       subtitle = "Relationship between robbery and unlawful acts",
       x = "Robbery",
       y = "Unlawful Acts") +
  theme_solarized() +
  theme(axis.title = element_text())
```

The plot shows that robbery and unlawful acts involving controlled drugs or precursors have a weak positive correlation.


# Task 3


```{r}
# Plotting overall offences on the European map 
# to quickly identify the high offending countries
# We will combine ggplot2's map_data with our raw EurostatCrime2019 data

# Loading packages
library(ggplot2)              
library(tidyverse)
library(cowplot)

# Using the original data set to get a complete list of the European countries
# We are not using row.names = 1 in this instance 
# as we need the countries as a column to preform a join with ggplot2's map data
CreativeStats <- read.csv(file = "C:\\Users\\jason\\OneDrive - University College Dublin\\Documents\\MSc Computer Science\\Final Semester\\STAT40730 - Data Prog with R\\Assignment 1\\EurostatCrime2019.csv", fileEncoding = "UTF-8-BOM")

# Creating a column containing the overall record of offences for each country
CreativeStats$offences <- rowSums(Filter(is.numeric, CreativeStats), na.rm = TRUE)

# Extracting two columns to a new database 
# which include the country and overall offences
EuropeMap <- CreativeStats[ , c("X", "offences")]  
head(EuropeMap)

# Changing the column name to match ggplot2's in order to preform a left join
colnames(EuropeMap)[colnames(EuropeMap) == "X"] <- "region"

# Renaming mismatching Eurostat country 
# names to align with ggplot2's country names
EuropeMap$region[EuropeMap$region 
                 %in% c("Czechia", 
                        "England and Wales", 
                        "Germany (until 1990 former territory of the FRG)",
                        "Kosovo (under United Nations 
                        Security Council Resolution 1244/99)", 
                        "Northern Ireland (UK)")] <- 
  c("Czech Republic", 
    "UK", "Germany", "Kosovo", "Northern Ireland")

head(EuropeMap)

# Creating a new dataframe to store ggplot2's mapdata
mapdata <- map_data("world")
View(mapdata)

# Left joining our crime data on the column region
mapdata <- left_join(mapdata, EuropeMap, by="region")
View(mapdata)

# Filtering NA's and storing in new dataframe
mapdata1<-mapdata %>% filter(!is.na(mapdata$offences))
View(mapdata1)

# Plotting our first map
map1<-ggplot(mapdata1, aes( x = long, y = lat, group=group)) +
  geom_polygon(aes(fill = offences), color = "black")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        rect = element_blank())
map1

# Plotting our second map
map2 <- map1 + scale_fill_gradient(name = "offences", 
                                   low = "yellow", 
                                   high =  "red", 
                                   na.value = "blue")+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        rect = element_blank())
map2

```

It is interesting to see that overall offences do not seem to correlate with countries that are considered less wealthy as one might expect. For example, Sweden is listed as having the highest overall offences but is listed as 5th on Europe's GDP per capita in 2019 according to [Eurostat's research](https://ec.europa.eu/eurostat/web/products-datasets/-/sdg_08_10). Similarly, Denmark is ranked second in overall offences and is listed as the 4th highest GDP per capita according to the same source.

On the other end of the scale, many eastern European countries such as Albania, Montenegro & Kosovo have some of the lowest offences and while being on the lower end of the GDP per capita scale for Europe.

This goes against the preconception that offences are linked to wealth. This is immediately apparent from the two plots where we can see that richer Scandinavian countries are in the higher end of the offences scale whereas less wealthy eastern European countries are in the lower end. This is a fascinating observation that we have not yet discovered in the dataset thus far.



